maxent model avec xvalidation (year-based test training)

### 2019-05-02 src/maxent_avec_xvalidation.Rmd
ok models good... about to add writebackgroundpredictions for cbi analysis...

```{r pre-install & load packages, include=FALSE}
options(java.parameters = "-Xmx1g" )#needs to run before dismo loaded - boosts memory for java to 1g
packages <- c("dismo", "stringr", "rJava", "raster", "ecospat", "bossMaps")
source("../src/ipak.R")
ipak(packages)
```

@@

#binary maps - individual layers

there seem to be two primary ways to create binary maps. the most common for conservation purposes seems to be to use maxsumspensspec values, however MPA also comes recommended (but is def less common). probably best to use cumulative for both (merrow 2013)

@@

for some reason the cumulative output comes empty... using bossMaps
```{r raw to cumulative}
rawasc_list <- list.files(paste0("../output/maxent/predictions/", mthna), pattern = "^raw_.*.asc", full.names = TRUE) 
for (raw in 1:length(rawasc_list)){
  asc <- raster(rawasc_list[raw])
  ascname <- names(asc)
  ascname <- str_sub(ascname, start = 5)
  cumasc <- cumulative(asc, threshold = NULL)
  writeRaster(cumasc, paste0("../output/maxent/predictions/", mthna, "/", "cum_", ascname, ".asc"), overwrite = TRUE)
}
```

```{r set up threshold binary function}
binmpa <- function(x) {
  ifelse(x <=  mpa_thr, 0,
  ifelse(x >  mpa_thr, 1, NA))
  }
```
cum_mpathr <- calc(cumasc, fun = rc)
plot(cum_mpathr)
tenper_thr <- 20.338725

```{r binary maps on individual layers}
cumasc_list <- list.files(paste0("../output/maxent/predictions/", mthna), pattern = "^cum_.*.asc", full.names = TRUE)

for (modtime in 1:nrow(modelmonth)){
  timeslice <- modelmonth[modtime]
  print(timeslice)
  yr <- str_sub(timeslice, -4)
  print(yr)
  thresholds <- read.csv(paste0("../output/maxent/models/", timeslice, "/AICforMaxentReg.csv"), header = TRUE)
  mpa_thr <- thresholds$mpa_0.9_cum
  print(mpa_thr)
  for (cum in 1:length(cumasc_list)){
    cumascyr <- str_sub(cumasc_list[cum], start = 40, end = 43) #selects the year from the file name
    if(yr == cumascyr){
      cumasc <- raster(cumasc_list[cum])
      ascname <- names(cumasc)
      print(ascname)
      cum_mpathr <- calc(cumasc, fun = binmpa)
      mpa_plot <- plot(cum_mpathr, png(paste0("../output/maxent/predictions/", mthna, "/", "binary_mpa_", ascname, ".png")))
      dev.off() # stops automatic saving of the plot to a png
      writeRaster(cum_mpathr, paste0("../output/maxent/predictions/", mthna, "/", "binary_mpa_", ascname, ".asc"), overwrite = TRUE)
    }
  }
}

@@

#model averageing

@@

```{r avg asc layers cloglog}
asc_list <- list.files(paste0("../output/maxent/predictions/", mthna, "/"), pattern = "^cloglog.*asc", full.names = TRUE)
depthlayers <- data.frame(cbind(unique(stdep$ascdepths)))
names(depthlayers)[names(depthlayers) == "cbind.unique.stdep.ascdepths.."] <- "depth"
depthlayers$depth <- as.character(depthlayers$depth)

@@

for (d in 1:nrow(depthlayers)){
  dep <- depthlayers$depth[d]
  for (a in 1:length(asc_list)){
    asc <- asc_list[a]
    if (grepl(dep, asc)){
      depthlayers[d , paste0("a", colno)] <- asc
        colno <- colno+1
    }
  }
}
write.csv(depthlayers, (paste0("../output/maxent/predictions/", mthna, "/cloglog_asclayerbydepth.csv")), row.names = TRUE)

@@

  stkasc <- stack(asc2003, asc2004, asc2005, asc2006)
  avg <- calc(stkasc, fun = mean)

@@

```{r avg asc layers cumulative}
asc_list <- list.files(paste0("../output/maxent/predictions/", mthna, "/"), pattern = "^cum.*asc", full.names = TRUE)
depthlayers <- data.frame(cbind(unique(stdep$ascdepths)))
names(depthlayers)[names(depthlayers) == "cbind.unique.stdep.ascdepths.."] <- "depth"
depthlayers$depth <- as.character(depthlayers$depth)

for (d in 1:nrow(depthlayers)){
  dep <- depthlayers$depth[d]
  for (a in 1:length(asc_list)){
    asc <- asc_list[a]
    if (grepl(dep, asc)){
      depthlayers[d , paste0("a", colno)] <- asc
        colno <- colno+1
    }
  }
}
write.csv(depthlayers, (paste0("../output/maxent/predictions/", mthna, "/cum_asclayerbydepth.csv")), row.names = TRUE)

  stkasc <- stack(asc2003, asc2004, asc2005, asc2006)
  avg <- calc(stkasc, fun = mean)

  writeRaster(avg, paste0("../output/maxent/predictions/", mthna, "/avg_cum_", asclaydep, ".asc"), overwrite=TRUE)
  avgplot <- plot(avg, png(paste0("../output/maxent/predictions/", mthna, "/avg_cum_", asclaydep, ".png"))) 
  dev.off()
}

@@

```{r binary maps on averaged rasters}
avgcumasc_list <- list.files(paste0("../output/maxent/predictions/", mthna), pattern = "^avg_cum.*.asc", full.names = TRUE)
thresholds <- read.csv(paste0("../output/maxent/models/", mthna, "avgAIC.csv"), header = TRUE)
thresholds <- tail(thresholds, 1) #subsset the last row
mpa_thr <- thresholds$mpa_0.9_cum
#print(mpa_thr)

for (avgcum in 1:length(avgcumasc_list)){
    cumasc <- raster(avgcumasc_list[avgcum])
    ascname <- names(cumasc)
    #print(ascname)
    cum_mpathr <- calc(cumasc, fun = binmpa)
    mpa_plot <- plot(cum_mpathr, png(paste0("../output/maxent/predictions/", mthna, "/", "binary_mpa_", ascname, ".png")))
    dev.off() # stops automatic saving of the plot to a png
    writeRaster(cum_mpathr, paste0("../output/maxent/predictions/", mthna, "/", "binary_mpa_", ascname, ".asc"), overwrite = TRUE, datatype="LOG1S")
}

@@

```{r}
test <- cellStats(binaryavgstack[[30]], sum)
test
```

>>>>

```{r pre-install & load packages, include=FALSE}
options(java.parameters = "-Xmx1g" )#needs to run before dismo loaded - boosts memory for java to 1g
packages <- c("dismo", "stringr", "rJava", "raster", "ecospat", "bossMaps", "lattice", "tidyr")
source("../src/ipak.R")
ipak(packages)
```

@@

#binary maps - individual year layers

there seem to be two primary ways to create binary maps. the most common for conservation purposes seems to be to use maxsumspensspec values, however MPA also comes recommended (but is def less common).

@@

```{r set up threshold binary function}
binmpa <- function(x) {
  ifelse(x <=  mpa_thr, 0,
  ifelse(x >  mpa_thr, 1, NA))
  }
```

```{r binary maps on individual layers cloglog}
clogasc_list <- list.files(paste0("../output/maxent/predictions/", mthna), pattern = "^cloglog_.*.asc", full.names = TRUE)

for (modtime in 1:nrow(modelmonth)){
  timeslice <- modelmonth[modtime]
  print(timeslice)
  yr <- str_sub(timeslice, -4)
  print(yr)
  thresholds <- read.csv(paste0("../output/maxent/models/", timeslice, "/AICforMaxentReg.csv"), header = TRUE)
  mpa_thr <- thresholds$mpa_0.9_clog
  print(mpa_thr)
  for (cum in 1:length(clogasc_list)){
    cumascyr <- str_sub(clogasc_list[cum], start = 44, end = 47) #selects the year from the file name
    if(yr == cumascyr){
      cumasc <- raster(clogasc_list[cum])
      ascname <- names(cumasc)
      print(ascname)
      cum_mpathr <- calc(cumasc, fun = binmpa)
      writeRaster(cum_mpathr, paste0("../output/maxent/predictions/", mthna,  "/binary_", ascname, ".asc"), overwrite = TRUE)
      plot(cum_mpathr, png(paste0("../output/maxent/predictions/", mthna, "/binary_", ascname, ".png")), overwrite = TRUE)
      dev.off() # stops automatic saving of the plot to a png
    }
  }
}
```

now for each year create a 2-d representation of suitable habitat

need to stack the rasters by year
run cellStats (sum)
save the output as they year

```{r binary map per year 2D representation}
binaryclog_list <- list.files(paste0("../output/maxent/predictions/", mthna), pattern = "^binary_.*.asc", full.names = TRUE)
yrlayer <- data.frame(matrix(ncol = 2, nrow = length(binaryclog_list)))
colnames(yrlayer) <- c("layer", "year")

for (yrlyr in 1:length(binaryclog_list)){
  yearoflyr <- str_sub(binaryclog_list[yrlyr], start = 51, end = 54) #selects the year from the file name
  yrlayer$year[yrlyr] <- yearoflyr
  yrlayer$layer[yrlyr] <- binaryclog_list[yrlyr]
}

y_list <- split(yrlayer, yrlayer$year)
noyrs <- unique(yrlayer$year)

for (binyr in (1:length(y_list))){
  binstack <- stack(y_list[[binyr]]$layer)
  year <- y_list[[binyr]]$year[1]
  stacksum <- sum(binstack, na.rm = TRUE)
  writeRaster(stacksum, paste0("../output/maxent/predictions/", mthna, "/binary_2d_", year, ".asc"))
  plot(stacksum, png(paste0("../output/maxent/predictions/", mthna, "/binary_2d_", year, ".png")))
  dev.off()
}

@@

#using averageing

@@

```{r avg asc layers cloglog}
clogasc_list <- list.files(paste0("../output/maxent/predictions/", mthna), pattern = "^cloglog_.*.asc", full.names = TRUE)
stdep <- read.csv(paste0("../output/maxent/predictions/", mthna, "/stack_depth.csv"), header = TRUE)
depthlayers <- data.frame(cbind(unique(stdep$ascdepths)))
names(depthlayers)[names(depthlayers) == "cbind.unique.stdep.ascdepths.."] <- "depth"
depthlayers$depth <- as.character(depthlayers$depth)

@@

for (d in 1:nrow(depthlayers)){
  dep <- depthlayers$depth[d]
  for (a in 1:length(clogasc_list)){
    asc <- clogasc_list[a]
    if (grepl(dep, asc)){
      depthlayers[d , paste0("a", colno)] <- asc
        colno <- colno+1
    }
  }
}
write.csv(depthlayers, (paste0("../output/maxent/predictions/", mthna, "/cloglog_asclayerbydepth.csv")), row.names = TRUE)

@@

  stkasc <- stack(asc2003, asc2004, asc2005, asc2006)
  avg <- mean(stkasc, na.rm = TRUE)

@@

```{r binary maps on averaged depth layer}
avgasc_list <- list.files(paste0("../output/maxent/predictions/", mthna), pattern = "^avg_cloglog_.*.asc", full.names = TRUE)
thresholds <- read.csv(paste0("../output/maxent/models/", mthna, "avgAIC.csv"), header = TRUE)
thresholds <- tail(thresholds, 1) #subset the last row
mpa_thr <- thresholds$mpa_0.9_clog

for (avg in 1:length(avgasc_list)){
    cumasc <- raster(avgasc_list[avg])
    ascname <- names(cumasc)
    mpathr <- calc(cumasc, fun = binmpa)
    mpa_plot <- plot(mpathr, png(paste0("../output/maxent/predictions/", mthna, "/binary_", ascname, ".png")))
    dev.off() # stops automatic saving of the plot to a png
    writeRaster(mpathr, paste0("../output/maxent/predictions/", mthna, "/binary_", ascname, ".asc"), overwrite = TRUE)
}

@@

```{r binary map averaged 2D representation}
binaryclog_list <- list.files(paste0("../output/maxent/predictions/", mthna), pattern = "^binary_avg_cloglog.*.asc", full.names = TRUE)

binstack <- stack(binaryclog_list)
stacksum <- sum(binstack, na.rm = TRUE)
writeRaster(stacksum, paste0("../output/maxent/predictions/", mthna, "/binary_avg_2d.asc"))
plot(stacksum, png(paste0("../output/maxent/predictions/", mthna, "/binary_avg_2d.png")))
dev.off()

@@

```{r calc no presence cells per layer (avg)} 
ascdepth_list <- list.files(paste0("../output/maxent/predictions/", mthna), pattern = "^binary_avg.*asc", full.names = TRUE)
binaryavgstack <- stack(ascdepth_list)
stackstat <- data.frame(matrix(ncol = 2, nrow = nlayers(binaryavgstack)))
stackstat[ ,1] <- c(1:49)
colnames(stackstat) <- c("layer", "cell_sum")

for (lyr in 1:nlayers(binaryavgstack)){
  lyrsum <- cellStats(binaryavgstack[[lyr]], sum)
  stackstat$cell_sum[lyr] <- lyrsum
}

write.csv(stackstat, paste0("../output/maxent/predictions/", mthna, "/cellcount_bydepth_avg.csv"), row.names = FALSE)
```

a crude plot but ok to show slr

```{r graph by depth (avg)}
bydepth <- barchart(layer ~ cell_sum, data = stackstat, ylim = c(49, 1), xlim = c(0, 300))
trellis.device(device="png", filename = paste0("../output/maxent/predictions/", mthna, "/depthplot_avg.png"))
print(bydepth)
dev.off()
```
